# Clustering and classification

&nbsp;

## Brief description of the dataset

&nbsp;

In this chapter the purpose is to get familiar with clustering and classification. I'm going to use a dataset called **Boston** which contains information collected by the U.S Census Service concerning housing in the area of Boston Mass. The data was originally published by Harrison, D. and Rubinfeld, D.L. "Hedonic prices and the demand for clean air", J. Environ. Economics & Management, vol.5, 81-102, 1978. You can find some additional information [here](http://www.clemson.edu/economics/faculty/wilson/R-tutorial/analyzing_data.html) and [here](https://archive.ics.uci.edu/ml/machine-learning-databases/housing/).

This dataset is actually already loaded into R thus let's take a look at some details.

&nbsp;

<p class="text-danger">**DATA VARIABLES**</p>

* crim - per capita crime rate by town
* zn - proportion of residential land zoned for lots over 25,000 sq.ft.
* indus - proportion of non-retail business acres per town
* chas - Charles River dummy variable (= 1 if tract bounds river; 0 otherwise) 
* nox - nitrogen oxides concentration (parts per 10 million)
* rm - average number of rooms per dwelling
* age - proportion of owner-occupied units built prior to 1940
* dis - weighted mean of distances to five Boston employment centres
* rad - index of accessibility to radial highways
* tax - full-value property-tax rate per \$10,000
* ptratio - pupil-teacher ratio by town
* black - 1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town
* lstat - lower status of the population (percent)
* medv - median value of owner-occupied homes in \$1000s


&nbsp;

<p class="text-danger">**STRUCTURE OFTHE BOSTON DATASET**</p>

From the structure table you can notice that there are 506 observation and 14 variables including 13 continuous variables and 1 binary-valued variable "chas". 

&nbsp;

```{r}
library(MASS)
data("Boston")
str(Boston)
```

&nbsp;

## Summaries and graphs

&nbsp;

<p class="text-success">**SUMMARY TABLE**</p>



```{r}
summary(Boston)
```

&nbsp;

```{r, fig.width=8, fig.height=8, message=FALSE, warning=FALSE}
library(ggplot2); library(gridExtra);library(dygraphs)

# INITIALIZING THE PLOTS

# Histogram of crim
hcrim <- qplot(Boston$crim, geom = "histogram", binwidth = 6, main = "Crime rate by town", xlab = "crime rate", col = I("grey"), fill = I("chartreuse3"))

# Histogram of zn
hzn <- qplot(Boston$zn, geom = "histogram", binwidth = 6, main = "Prop. of residential land zoned", xlab = "zn", col = I("grey"), fill = I("orange"))

# Histogram of indus
hindus <- qplot(Boston$indus, geom = "histogram", binwidth = 3, main = "Prop. of non-retail business", xlab = "indus", col = I("grey"), fill = I("mediumorchid"))

hnox <- qplot(Boston$nox, geom = "histogram", binwidth = 0.02, main = "Nitrogen oxides concentration", xlab = "nox", col = I("grey"), fill = I("blue"))


# Combining the plots using the function grid.arrange() [in gridExtra] 
  
# Additional information from (http://www.sthda.com/english/wiki/ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page-r-software-and-data-visualization)
grid.arrange(hcrim, hzn, hindus, hnox, ncol=2, nrow =2)
```



&nbsp;

```{r, message=FALSE, warning=FALSE}
library(ggplot2); library(gridExtra)

# INITIALIZING THE PLOTS

hrm <- qplot(Boston$rm, geom = "histogram", binwidth = 1, main = "average number of rooms per dwelling", xlab = "rm", col = I("grey"), fill = I("yellow"))
hrm

hage <- qplot(Boston$age, geom = "histogram", binwidth = 10, main = "prop. of owner-occupied units built prior to 1940", xlab = "age", col = I("grey"), fill = I("brown"))
hage


# Combining the plots using the function grid.arrange() [in gridExtra] 

grid.arrange(hrm, hage, ncol=2, nrow =2)

```


&nbsp;




```{r, message=FALSE, warning=FALSE}
library(tidyverse)
cor_matrix <- cor(Boston) %>% round(digits = 2)
cor_matrix
```

&nbsp;


ABOUT THE CORRELATION MATRIX

Positive correlations are displayed in blue and negative correlations in red color. Color intensity and the size of the circle are proportional to the correlation coefficients. The legend color shows the correlation coefficients and the corresponding colors.


```{r, fig.width=10, fig.height=10}
library(corrplot); library(dygraphs)
corrplot(cor_matrix, method = "circle", main = "Correlation matrix of Boston variables", tl.cex = 1, addCoef.col = "black")

```


&nbsp;




&nbsp;
