# Logistic Regression

&nbsp;

SOMETHING HERE, WHAT IM GOING TO DO IN THIS CHAPTER

&nbsp;


## Data description 

&nbsp;

I'm going to use a dataset of **Student alcohol consumption** from UCI Machine Learning Repository. The purpose of the original data have been to predict secondary school student alcohol comsumption in Portugal. This material contains two separate datasets of math and Portuguese language courses but to note that the questionnaires have been identical. There are also students who have answered both questionnaires. 

&nbsp;

=====================================

**You can find more detailed information about those two datasets**   [here](https://archive.ics.uci.edu/ml/datasets/STUDENT+ALCOHOL+CONSUMPTION).


**Data reference**: Using Data Mining To Predict Secondary School Student Alcohol Consumption. Fabio Pagnotta, Hossain Mohammad Amran. Department of Computer Science,University of Camerino.

======================================

&nbsp;

**In this analysis** the purpose is to use combined datset where we have those students who have answered these two questionnaires i.e. students who have participated both in the math and the Portuguese language courses. To combine the two datasets we have utilized some of the backround questions including: 

* school (student's school: GB - Gabriel Pereira or MS - Mousinho da Silveira)
* sex (F - female or M - male)
* age (student's age:  from 15 to 22)
* address (U - urban or R - rural)
* famsize (family size: LE3 - less or equal to 3 or GT3 - greater than 3)
* Pstatus (parent's cohabitation status: T - living together or A - apart)
* Medu (mother's education:  0 - none, 1 - primary education (4th grade), 2 - 5th to 9th grade, 3 - secondary education or 4 - higher education)
* Fedu (father's education: 0 - none, 1 - primary education (4th grade), 2 - 5th to 9th grade, 3 - secondary education or 4 - higher education)
* Mjob (mother's job: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')
* Fjob (father's job: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')
* reason (reason to choose this school: close to 'home', school 'reputation', 'course' preference or 'other)
* nursery (attended nursery school: yes or no)
* internet (Internet access at home: yes or no)

&nbsp;

======================================

**You can find the dataset which is utilized in this analysis** [here](http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/alc.txt).

======================================

&nbsp;

**Underneath you can see all the variables of the combined dataset** consisting 382 observations and 35 variables. 

The following adjustments have also been made: The variables not used for joining the two data have been combined by averaging (including the grade variables). "alc_use"" is the average of "Dalc" (workday alcohol consumption) and "Walc" (weekend alcohol consumption). "high_use"" is TRUE if "alc_use" is higher than 2 and the scale ranges between 1 and 5 (from 1 - very low to 5 - very high). 



&nbsp;

```{r}
joined_alcohol <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/alc.txt", sep = ",", header = TRUE)
```

&nbsp;

<p class="text-danger">**NAMES OF DATA VARIABLES**</p>


```{r}
colnames(joined_alcohol)
```

&nbsp;

**Underneath you can take a closer look at the content of the data variables.** The variables which i haven't introduce yet includes:

* guardian (student's guardian: 'mother', 'father' or 'other')
* traveltime (home to school travel time: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour) 
* studytime (weekly study time: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours) 
* failures (number of past class failures: n if 1 <= n < 3, else 4) 
* schoolsup (extra educational support: yes or no)
* famsup (family educational support: yes or no) 
* paid (extra paid classes within the course subject (Math or Portuguese): yes or no) 
* activities (extra-curricular activities: yes or no) 
* higher (wants to take higher education: yes or no) 
* romantic (with a romantic relationship: yes or no) 
* famrel (quality of family relationships: from 1 - very bad to 5 - excellent) 
* freetime (free time after school: from 1 - very low to 5 - very high)
* goout (going out with friends: from 1 - very low to 5 - very high) 
* Dalc (workday alcohol consumption: from 1 - very low to 5 - very high) 
* Walc (weekend alcohol consumption: from 1 - very low to 5 - very high) 
* health (current health status: from 1 - very bad to 5 - very good) 
* absences (number of school absences: from 0 to 93) 
* G1 (first period grade: from 0 to 20) 
* G2 (second period grade: from 0 to 20) 
* G3 (final grade: from 0 to 20, output target) 

&nbsp;

<p class="text-danger">**STRUCTURE OF THE DATA VARIABLES**</p>

```{r}
str(joined_alcohol)
```


&nbsp;

## My initial hypotheses 

&nbsp;

The purpose of this analysis is to study the relationships between high/low alcohol consumption and 4 variables of my own interest in the data. I think there are many interesting variables which might have the relationship with the alcohol consumption. 

&nbsp;

<p class="text-success">**HYPOTHESIS 1**</p>

First variable will be **sex** i.e. female or male. I believe the gender has a relationship with alcohol consumption but it is impossible to say which gender type has a bigger probability to be a "high user" or "low user". One of the reasons are that I'm unfamiliar with the culture of Portugal and their habits of alcohol consumption. 

&nbsp;

<p class="text-success">**HYPOTHESIS 2**</p>

The second interesting variable will be the weekly **studytime** which has the possible outcomes of 1 (less than 2 hours), 2 (2 to 5 hours), 3 (5 to 10 hours) or 4 (more than 10 hours). 

I believe that the weekly studytime and alcohol consumption have a relationship in the way that the less a student studies the bigger is the probability to be an alcohol "high user" and vice versa. 

&nbsp;

<p class="text-success">**HYPOTHESIS 3**</p>

The third variable which I'm going to choose is **goout** as going out with friends. This variable has the possible values from 1 (very low) to 5 (very high). 

The variable of going out with friends doesn't tell us which kind of activities it consists but I believe that the more one goes out the bigger probability is to use alcohol. That is the more the student goes out the greater might be the probability to be an alcohol "high user" and vice versa. 

&nbsp;

<p class="text-success">**HYPOTHESIS 4**</p>

The last variable which interest me is **famrel** as quality of family relationships which ranges from 1 (very bad) to 5 (excellent).

I believe there can be a relationship between family relationships and alcohol consumption in such a way that the worse these relationships are the bigger is the probability to be a "high user" of alcohol. Of course in a bigger picture the other relevant aspect could be the overall importance of these family relationships which can define if these relationships affect to the student's alcohol consumption at all.   


&nbsp;


## Relationships between the variables in terms of my initial hypotheses

&nbsp;

In this section I'm going to explore the distributions of the variables which were introduced in the previous chapter. The second purpose is to explore the possible relationships between alcohol consumption and the chosen variables. 

&nbsp;

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyr); library(dplyr); library(ggplot2); library(knitr)
```

&nbsp;

<p class="text-warning">**SEX AND ALCOHOL CONSUMPTION**.</p>


As you can see from the table below there are 198 females and 184 males in the dataset thus the groups seem to be nearly same size. Furthermore there are 382 students in the dataset.

&nbsp;

```{r}
sex <- joined_alcohol$sex
alc_use <- joined_alcohol$alc_use
high_use <- joined_alcohol$high_use
joined_alcohol %>% group_by(sex) %>% summarise(count = n())
```

&nbsp;

Im writing

&nbsp;

```{r}
joined_alcohol %>% group_by(sex, high_use) %>% summarise(count = n())
```

&nbsp;


<p class="text-warning">**WEEKLY STUDYTIME AND ALCOHOL CONSUMPTION**.</p>

Writing here

&nbsp;

```{r}
studytime <- joined_alcohol$studytime
joined_alcohol %>% group_by(studytime) %>% summarise(count = n())
```


<p class="text-warning">**GOING OUT WITH FRIENDS AND ALCOHOL CONSUMPTION**.</p>

Writing here

&nbsp;

```{r}
goout <- joined_alcohol$goout
joined_alcohol %>% group_by(goout) %>% summarise(count = n())
```


<p class="text-warning">**FAMILY RELATIONSHIPS AND ALCOHOL CONSUMPTION**.</p>

Writing here

&nbsp;

```{r}
famrel <- joined_alcohol$famrel
joined_alcohol %>% group_by(famrel) %>% summarise(count = n())
```





===================================

Numerically and graphically explore the distributions of your chosen variables and their relationships with alcohol consumption (use for example cross-tabulations, bar plots and box plots). Comment on your findings and compare the results of your exploration to your previously stated hypotheses. (5 points)

===================================

&nbsp;


