# Regression and model validation

In this section I'm going to analyze the data set of students who participated to the course called **Introduction to Social Statistics** at Helsinki University in the fall 2014. These 183 students were asked to participate to the survey about their Learning Approaches consisting deep, strategic and surface orientated learning. In addition the students were asked to evaluate their attitude towars statistics.

You can find some basic information about the variables [here](http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-meta.txt)  


*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using  


## Introducing the data set

I'm going to use partly the above mentioned data which still contains the different learning orientations and attitudes towars statistics. In addition the data variables consist students' age, gender and course exam points.  

***

**Here you can find the data set what I'm going to use in this analysis**

```{r}
students2014 <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt", header = TRUE, sep = ",")
```

***
**Then I explore what kind of material the students' data set contains**

As you can see below, there is 166 observations i.e. students and 7 variables. The variables contains the age and gender of the students. The genders can be recogniced **F** as woman and **M** as man. 

The attitudes and learning orientations were measured on the Likert scale from 1 to 5. The attitude towards statistics is a sum of 10 questions, the deep orientation is 12, the strategic is 8 and the surface orientation is a sum of 12 question. Last but not least the data contains the exam points of the course.

Under the structure table you can see there is no missing data (False = 1162). 

***

```{r}
str(students2014)
table(is.na(students2014))
```

***

## Summaries and graphs 

Underneath you can see the data summary. It tells there are 110 (66 %) women and 56 (34 %) men. Furthermore the students' age range between 17 and 55 where the mean age is around 25 and median is 22. KAIKKI MUUTTUJAT PITÄÄ KUVAILLA! 

***

```{r}
summary(students2014)
```


```{r}
library(ggplot2)
library(GGally)
```

***

**First you can see scatter plot matrix with every pair in the data besides gender**

TÄHÄN VIELÄ TEKSTIÄ


```{r}
pairs(students2014[-1], col = students2014$gender)
```

***

**Next I take couple examples to explore more closely some scatter plots**

TÄHÄN TEKSTIÄ YKSITTÄISISTÄ SCATTER PLOTEISTA

```{r}
attitude <- students2014$attitude
points <- students2014$points
gender <- students2014$gender
ggplot(students2014, aes(x = attitude, y = points, col = gender)) + geom_point() + xlab("Attitudes towards statistics") + ylab("Exam points") + ggtitle("Attitudes vs. Exam points")
ggplot()
```






