# Clustering and classification

&nbsp;

## Brief description of the dataset

&nbsp;

In this chapter the purpose is to get familiar with clustering and classification. I'm going to use a dataset called **Boston** which contains information collected by the U.S Census Service concerning housing in the area of Boston Mass. The data was originally published by Harrison, D. and Rubinfeld, D.L. `Hedonic prices and the demand for clean air', J. Environ. Economics & Management, vol.5, 81-102, 1978. You can find some additional information [here](http://www.clemson.edu/economics/faculty/wilson/R-tutorial/analyzing_data.html)

This dataset is actually already loaded into R thus let's take a look at some details.

&nbsp;

<p class="text-danger">**DATA VARIABLES**</p>

* crim - per capita crime rate by town
* zn - proportion of residential land zoned for lots over 25,000 sq.ft.
* indus - proportion of non-retail business acres per town
* chas - Charles River dummy variable (= 1 if tract bounds river; 0 otherwise) 
* nox - nitrogen oxides concentration (parts per 10 million)
* rm - average number of rooms per dwelling
* age - proportion of owner-occupied units built prior to 1940
* dis - weighted mean of distances to five Boston employment centres
* rad - index of accessibility to radial highways
* tax - full-value property-tax rate per \$10,000
* ptratio - pupil-teacher ratio by town
* black - 1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town
* lstat - lower status of the population (percent)
* medv - median value of owner-occupied homes in \$1000s


&nbsp;

<p class="text-danger">**STRUCTURE OFTHE BOSTON DATASET**</p>

From the structure table you can notice that there are 506 observation and 14 variables.

&nbsp;

```{r}
library(MASS)
data("Boston")
str(Boston)
```

&nbsp;

## Summaries and graphs

&nbsp;

<p class="text-success">**SUMMARY TABLE**</p>



```{r}
summary(Boston)
```

&nbsp;

xx



```{r, message=FALSE, warning=FALSE}
install.packages("corrplot")
```

```{r, message=FALSE, warning=FALSE}
library(corrplot); library(tidyverse)
cor_matrix <- cor(Boston) %>% round(digits = 2)
cor_matrix
```

&nbsp;


&nbsp;

