# Dimensionality reduction techniques 

&nbsp;

In this chapter the purpose is to get familiar with some of the Dimensionality Reduction Techniques including Principal Component Analysis (PCA) and Multiple Correspondece Analysis (MCA). 

Principal Component Analysis is way of identifying patterns in data, and expressing the data in such a way as to highlight their similarities and differences. PCA is a procedure After founding the patterns PCA has the tools of compressing the data by reducing the number of dimensions, without much loss of information. [Additional info of PCA](https://en.wikipedia.org/wiki/Principal_component_analysis)

Multiple Correspondence Analysis is used for the same purpose as PCA but it can be utilized if the data contains categorical variables. [Additional info of MCA](https://en.wikipedia.org/wiki/Multiple_correspondence_analysis)


&nbsp;

## Brief description of the dataset


&nbsp;

In this chapter I will be working with the **Data of Human Development approach** adopted from [United Nations Development programme, UNDP](http://hdr.undp.org/en/humandev). This programme has explored human development around the world with specific indices in each year since 1990, [Country List in 2015](http://hdr.undp.org/en/countries). The specified indices are: 

1. the Human Development Index, HDI
2. the Inequality-Adjusted Human Development Index, IHDI
3. the Gender Development Index, GDI
4. the Gender Inequality Index, GII
5. the Multidimensional Poverty Index, MPI

<br>

**I'm going to work with HDI and GII indices by combining the datasets and using specific variables including:**

1. country - technically a rowname
2. SFM_edu = SFedu / SMedu, where
    + SFedu - Females, Population with at least some secondary education (% ages 25 and older)
    + SMedu - Males, Population with at least some secondary education (% ages 25 and older)
3. FMlabour = Flabour / Mlabour, where
    + Mlabour - Males, Labour force participation rate (% ages 15 and older
    + Flabour - Females, Labour force participation rate (% ages 15 and older)
4. Eedu - Expected years of schooling per country
5. Elife - Expected life years at birth per country
6. GNIpc - Gross national income per capita per country in PPP, (https://en.wikipedia.org/wiki/Purchasing_power_parity)
7. mort_ratio - Maternal mortality ratio (deaths per 100,000 live births)
8. ad_birth - Adolescent birth rate (births per 1,000 women ages 15-19)
9. seats - Share of seats in parliament (% held by women)

<br>

**Note 1**, SFM_edu is a ratio of education where values less than 1 are signifying that males are more educated than females and values greater than 1 are implying that females are more educated than men.

**Note 2**, FMlabour is a ratio of labour force participation where values less than 1 are signifying that males are more participated than females and values greater than 1 are implying that females are more participated than men.

**Note 3**, variables from the Human Development Index data are Eedu, Elife and GNIpc while the other variables are from the Gender Inequality Index data.

&nbsp;

***

For additional information you can find [here](http://hdr.undp.org/en/composite/HDI) the Human Development Index data and its components.

For additional information you can find [here](http://hdr.undp.org/en/composite/GII) the Gender Inequality Index data and its components.

***


&nbsp;

<p class="text-danger">**COMBINED DATASET**</p>

You can find the combined dataset from my GutHub respitory: https://github.com/8475394/IODS-project/blob/master/data/human_1.txt

&nbsp;

```{r, message=FALSE, warning=FALSE}
library(tidyr); library(readr)
human_own <- read.table(file.path("C:/Users/heidi/Documents/YLIOPISTO/TILASTOTIEDE/INTRODUCTION TO OPEN DATA SCIENCE/IODS-project/data", "human_1.txt"), header = TRUE)
```

&nbsp;

<p class="text-danger">**STRUCTURE OF THE HUMAN DATASET**</p>

From the structure table you can notice that there are 155 observation and 8 variables. Note that the country is a rowname and it doesn't appear to variable list.

&nbsp;

```{r, message=FALSE, warning=FALSE}
str(human_own)
```

&nbsp;

## Summaries and graphs


&nbsp;


In this section I'm going to explore the distributions of data variables even further. First I show you a summary table which contains the distributions of each variable where the variables are divided into quartiles including the minimum and maximum values. 

After this I'll show you the histograms of these variables WHAT ...

At the end of this chapter I'll explore WHAT... 

&nbsp;

<p class="text-success">**SUMMARY TABLE OF DATA VARIABLES**</p>

You may see some intresting "phenomena" from the summary table. Both the ratio of education (SFM_edu) and labour force participation (FMlabour) suggest that males are more dominant in these fields. It seems that females are more equal in education than labour force participation. You can examine this visually from the histograms later in this chapter.

According to the [Technical notes](http://hdr.undp.org/sites/default/files/hdr2015_technical_notes.pdf) of the Human Development indices, the scale of expected years of schooling per country (Eedu) varies between 0 and 18 where the latter is equivalent to achieving a master's degree in most countries. However you can see that the actual maximum of the schooling distribution is a bit over 20. If I understood right the range of 0-18 is used when counting the overall Human Development Index. Nevertheless I think the average of expected schooling years is quite good even though these values don't tell us anything about specific countries.

I can't help myself to wonder the minimum value of Expected life years at birth per country (Elife). 49 years seem so low even though the mean and median values are much greater. 

I think the distribution of Gross National Income per capita PPP emphasizes the huge distinctions between countries and of their wealth. Note that GNI per capita in PPP measures purchasing power parity which is useful for making comparisons between countries, [PPP](https://en.wikipedia.org/wiki/Purchasing_power_parity). However, there is a but when measuring the Human Development Index. If you look at [the table of HDI](http://hdr.undp.org/en/composite/HDI)(as seen earlier in this chapter), you can perceive that the GNI is not necessary the most critical measure in terms of well-being. For example the GNI of Montenegro is lower than average but the Human Development Index in very high. Note also that this was kind of an additional information since the actual HD index is not a part of this examination.

Based on the technical notes the Maternal Mortality rate represent the maternal health, the higher the rate the poorer the maternal health. If you look at [the table of GII](http://hdr.undp.org/en/composite/GII)(as seen earlier in this chapter), you can perceive that the maximum value of the maternal mortality among countries with very high human development is 69. This is just to give you some understanding of this scale. 

Adolescent birth rate represents the rate of births per 1000 women between ages 15 and 19. I think this scale varies also a lot. However the distribution is skewed to the right (positive skewness) suggesting that there are more lower than higher values in this scale. 

The distribution of seats represents the share (%) of seats in parliament held by women in each country. Here you can see for example that it is not common to have more females than males in parliaments as the median and mean values are both close to 20 %. Also there are just few countries where the women are in majority.   



&nbsp;

```{r, results='asis'}
library(knitr)
kable(summary(human_own), format = "pandoc", digits = 2,  align="l")
```

&nbsp;

<p class="text-success">**HISTOGRAMS OF EDUCATION, LABOUR FORCE, YEARS OF SCHOOLING AND LIFESPAN**</p>


```{r, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
library(ggplot2); library(gridExtra);library(dygraphs)

# INITIALIZING THE PLOTS

h1 <- qplot(human_own$SFM_edu, geom = "histogram", binwidth = 0.1, main = "Population with at least sec. school", xlab = "ratio of education, female/male", ylab = "number of countries", col = I("grey"), fill = I("chartreuse3"))

h2 <- qplot(human_own$FMlabour, geom = "histogram", binwidth = 0.08, main = "Labour force participation rate", xlab = "ratio of labour force, female/male", ylab = "number of countries", col = I("grey"), fill = I("orange"))

h3 <- qplot(human_own$Eedu, geom = "histogram", binwidth = 1.2, main = "Expected years of schooling per country", xlab = "years per country", ylab = "number of countries", col = I("grey"), fill = I("mediumorchid"))

h4 <- qplot(human_own$Elife, geom = "histogram", binwidth = 4, main = "Expected life years at birth per country", xlab = "years per country", ylab = "number of countries", col = I("grey"), fill = I("blue"))

# Combining the plots using the function grid.arrange()

grid.arrange(h1, h2, h3, h4, ncol=2, nrow =2)
```


&nbsp;

<p class="text-success">**HISTOGRAMS OF GNI, MATERNAL MORTALITY, ADOLESCENT BIRTH RATE AND SHARE OF SEATS**</p>


```{r, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
library(ggplot2); library(gridExtra); library(dygraphs)

# INITIALIZING THE PLOTS

h5 <- qplot(human_own$GNIpc, geom = "histogram", binwidth = 10000, main = "Gross National Income, GNI", xlab = "GNI per capita PPP", ylab = "number of countries", col = I("grey"), fill = I("red"))

h6 <- qplot(human_own$mort_ratio, geom = "histogram", binwidth = 50, main = "Maternal mortality ratio", xlab = "deaths per 100 000 live births", ylab = "number of countries", col = I("grey"), fill = I("cornflowerblue"))

h7 <- qplot(human_own$ad_birth, geom = "histogram", binwidth = 10, main = "Adolescent birth rate", xlab = "births per 1000 women ages 15-19", ylab = "number of countries", col = I("grey"), fill = I("yellow"))

h8 <- qplot(human_own$seats, geom = "histogram", binwidth = 3, main = "Share of seats in parliament", xlab = "% held by women", ylab = "number of countries", col = I("grey"), fill = I("violetred2"))

# Combining the plots using the function grid.arrange()

grid.arrange(h5, h6, h7, h8, ncol=2, nrow =2)
```


&nbsp;


&nbsp;

<p class="text-success">**INITIAL RESULTS OF RELATIONSHIPS**</p>

&nbsp;


```{r, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
require(GGally); require(ggplot2)
g = ggpairs(human_own, lower = list(continuous = wrap("smooth", method = "lm", color="red")), upper = list(continuous = wrap("density", color = "blue")))
g
```



&nbsp;

&nbsp;

<p class="text-success">**CORRELATION MATRIX BETWEEN DATA VARIABLES**</p>



Here I'm going to explore the linear relationships between different variables more closely to get a better understanding of the data.

&nbsp;

```{r, message=FALSE, warning=FALSE}
library(tidyverse); library(knitr)
cor_matrix <- cor(human_own) %>% round(digits = 2)
kable(cor_matrix, format = "pandoc", digits = 2,  align="c")
```


&nbsp;

<p class="text-success">**PLOTTING THE CORRELATION MATRIX**</p>


```{r}
library(corrplot)
cor(human_own) %>% corrplot()
```


&nbsp;

