# Dimensionality reduction techniques 

&nbsp;

In this chapter the purpose is to get familiar with some Dimensionality Reduction Techniques including Principal Component Analysis (PCA) and Multiple Correspondece Analysis (MCA). 

Principal Component Analysis is a way of identifying patterns in data, and expressing the data in such a way as to highlight their similarities and differences. After founding the patterns PCA has the tools of compressing the data by reducing the number of dimensions, without much loss of information. [Additional info of PCA](https://en.wikipedia.org/wiki/Principal_component_analysis)

Multiple Correspondence Analysis is used for the same purposes as PCA but it can be utilized if the data contains categorical variables. [Additional info of MCA](https://en.wikipedia.org/wiki/Multiple_correspondence_analysis)


&nbsp;

## Brief description of the dataset


&nbsp;

In this chapter I will be working with the **Data of Human Development approach** adopted from [United Nations Development programme, UNDP](http://hdr.undp.org/en/humandev). This programme has explored human development around the world with specific indices in each year since 1990, [Country List in 2015](http://hdr.undp.org/en/countries). The specified indices are: 

1. the Human Development Index, HDI
2. the Inequality-Adjusted Human Development Index, IHDI
3. the Gender Development Index, GDI
4. the Gender Inequality Index, GII
5. the Multidimensional Poverty Index, MPI

<br>

**I'm going to work with HDI and GII indices by combining the datasets and using specific variables including:**

1. country - technically a rowname
2. SFM_edu = SFedu / SMedu, where
    + SFedu - Females, Population with at least some secondary education (% ages 25 and older)
    + SMedu - Males, Population with at least some secondary education (% ages 25 and older)
3. FMlabour = Flabour / Mlabour, where
    + Mlabour - Males, Labour force participation rate (% ages 15 and older
    + Flabour - Females, Labour force participation rate (% ages 15 and older)
4. Eedu - Expected years of schooling per country
5. Elife - Expected life years at birth per country
6. GNIpc - Gross national income per capita per country in PPP, (https://en.wikipedia.org/wiki/Purchasing_power_parity)
7. mort_ratio - Maternal mortality ratio (deaths per 100,000 live births)
8. ad_birth - Adolescent birth rate (births per 1,000 women ages 15-19)
9. seats - Share of seats in parliament (% held by women)

<br>

**Note 1**, SFM_edu is a ratio of education where values less than 1 are signifying that males are more educated than females and values greater than 1 are implying that females are more educated than men.

**Note 2**, FMlabour is a ratio of labour force participation where values less than 1 are signifying that males are more participated than females and values greater than 1 are implying that females are more participated than men.

**Note 3**, variables from the Human Development Index data are Eedu, Elife and GNIpc while the other variables are from the Gender Inequality Index data.

&nbsp;

***

For additional information you can find [here](http://hdr.undp.org/en/composite/HDI) the Human Development Index data and its components.

For additional information you can find [here](http://hdr.undp.org/en/composite/GII) the Gender Inequality Index data and its components.

***


&nbsp;

<p class="text-danger">**COMBINED DATASET**</p>

You can find the combined dataset from my GutHub respitory: https://github.com/8475394/IODS-project/blob/master/data/human_1.txt

&nbsp;

```{r, message=FALSE, warning=FALSE}
library(tidyr); library(readr)
human_own <- read.table(file.path("C:/Users/heidi/Documents/YLIOPISTO/TILASTOTIEDE/INTRODUCTION TO OPEN DATA SCIENCE/IODS-project/data", "human_1.txt"), header = TRUE)
```

&nbsp;

<p class="text-danger">**STRUCTURE OF THE HUMAN DATASET**</p>

From the structure table you can notice that there are 155 observation and 8 variables. Note that the country is actaully a rowname and it doesn't appear to variable list.

&nbsp;

```{r, message=FALSE, warning=FALSE}
str(human_own)
```

&nbsp;

## Summaries and graphs


&nbsp;


In this section I'm going to explore the distributions of data variables even further. First I show you a summary table which contains the distributions of each variable where the variables are divided into quartiles including the minimum and maximum values. 

After this I'll show you a set of histograms and explore the potential relationhips between each variable.


&nbsp;

<p class="text-success">**SUMMARY TABLE OF DATA VARIABLES**</p>

You may see some intresting "phenomena" from the summary table. Both the ratio of education (SFM_edu) and labour force participation (FMlabour) suggest that males are more dominant in these fields. It seems that females are more equal in education than labour force participation. You can examine this visually from the histograms later in this chapter.

According to the [Technical notes](http://hdr.undp.org/sites/default/files/hdr2015_technical_notes.pdf) of the Human Development indices, the scale of expected years of schooling per country (Eedu) varies between 0 and 18 where the latter is equivalent to achieving a master's degree in most countries. However you can see that the actual maximum of the schooling distribution is a bit over 20. If I understood right the range of 0-18 is used when counting the overall Human Development Index. Nevertheless I think the average of expected schooling years is quite good even though these values don't tell us anything about specific countries.

I can't help myself to wonder the minimum value of Expected life years at birth per country (Elife). 49 years seem so low even though the mean and median values are much greater. 

I think the distribution of Gross National Income per capita PPP emphasizes the huge distinctions between countries and of their wealth. Note that GNI per capita in PPP measures purchasing power parity which is useful for making comparisons between countries, [PPP](https://en.wikipedia.org/wiki/Purchasing_power_parity). However, there is a but when measuring the Human Development Index. If you look at [the table of HDI](http://hdr.undp.org/en/composite/HDI) (as seen earlier), you can perceive that the GNI is not necessary the most critical measure in terms of well-being. For example the GNI of Montenegro is lower than average but the Human Development Index is very high. Note also that this was kind of an additional information since the actual HD index is not a part of this examination.

Based on the technical notes the Maternal Mortality rate represent the maternal health, the higher the rate the poorer the maternal health. If you look at [the table of GII](http://hdr.undp.org/en/composite/GII) (as seen earlier), you can perceive that the maximum value of the maternal mortality among countries with very high human development is 69. This is just to give you some understanding of this scale. 

Adolescent birth rate represents the rate of births per 1000 women between ages between 15 and 19. I think this scale varies also a lot. However the distribution is skewed to the right (positive skewness) suggesting that there are more lower than higher values in this scale. 

The distribution of seats represents the share (%) of seats in parliament held by women in each country. Here you can see for example that it is not common to have more females than males in parliaments as the median and mean values are both close to 20 %. Also there are just few countries where the women are in majority.   



&nbsp;

```{r, results='asis'}
library(knitr)
kable(summary(human_own), format = "pandoc", digits = 2,  align="l")
```

&nbsp;

<p class="text-success">**HISTOGRAMS OF EDUCATION, LABOUR FORCE, YEARS OF SCHOOLING AND LIFESPAN**</p>

<br>

```{r, fig.height=8, fig.width=8, message=FALSE, warning=FALSE}
library(ggplot2); library(gridExtra);library(dygraphs)

# INITIALIZING THE PLOTS

h1 <- qplot(human_own$SFM_edu, geom = "histogram", binwidth = 0.1, main = "Population with at least sec. school", xlab = "ratio of education, female/male", ylab = "number of countries", col = I("grey"), fill = I("chartreuse3"))

h2 <- qplot(human_own$FMlabour, geom = "histogram", binwidth = 0.08, main = "Labour force participation rate", xlab = "ratio of labour force, female/male", ylab = "number of countries", col = I("grey"), fill = I("orange"))

h3 <- qplot(human_own$Eedu, geom = "histogram", binwidth = 1.2, main = "Expected years of schooling per country", xlab = "years per country", ylab = "number of countries", col = I("grey"), fill = I("mediumorchid"))

h4 <- qplot(human_own$Elife, geom = "histogram", binwidth = 4, main = "Expected life years at birth per country", xlab = "years per country", ylab = "number of countries", col = I("grey"), fill = I("blue"))

# Combining the plots using the function grid.arrange()

grid.arrange(h1, h2, h3, h4, ncol=2, nrow =2)
```


&nbsp;

<p class="text-success">**HISTOGRAMS OF GNI, MATERNAL MORTALITY, ADOLESCENT BIRTH RATE AND SHARE OF SEATS**</p>

<br>

```{r, fig.height=8, fig.width=8, message=FALSE, warning=FALSE}
library(ggplot2); library(gridExtra); library(dygraphs)

# INITIALIZING THE PLOTS

h5 <- qplot(human_own$GNIpc, geom = "histogram", binwidth = 10000, main = "Gross National Income, GNI", xlab = "GNI per capita PPP", ylab = "number of countries", col = I("grey"), fill = I("red"))

h6 <- qplot(human_own$mort_ratio, geom = "histogram", binwidth = 50, main = "Maternal mortality ratio", xlab = "deaths per 100 000 live births", ylab = "number of countries", col = I("grey"), fill = I("cornflowerblue"))

h7 <- qplot(human_own$ad_birth, geom = "histogram", binwidth = 10, main = "Adolescent birth rate", xlab = "births per 1000 women ages 15-19", ylab = "number of countries", col = I("grey"), fill = I("yellow"))

h8 <- qplot(human_own$seats, geom = "histogram", binwidth = 3, main = "Share of seats in parliament", xlab = "% held by women", ylab = "number of countries", col = I("grey"), fill = I("violetred2"))

# Combining the plots using the function grid.arrange()

grid.arrange(h5, h6, h7, h8, ncol=2, nrow =2)
```


&nbsp;


&nbsp;

<p class="text-success">**INITIAL RESULTS OF THE RELATIONSHIPS**</p>


Here you can have an initial look at the relationships between human data variables. I added a regression line into each scatterplot to better perceive the direction of the relationships. 


**NOTE**, According to James, Witten, Hastie & Tibshirani (2013, 375) these two-dimenional scatterplots are most likely not so informative when we want to explore the total information of the data. This is the case especially if we have a lot of variables. Futhermore these two-dimensional plots contain just a small fraction of the total information present in the dataset. 

However I will explore also the correlations between the variables before continuing the dimensionality reduction part. Correlation matrix and plot are seen after the initial results.


<br>

```{r, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
library(GGally); library(ggplot2)
g = ggpairs(human_own, lower = list(continuous = wrap("smooth", method = "lm", color="red")), upper = list(continuous = wrap("density", color = "blue")))
g
```



&nbsp;

&nbsp;

<p class="text-success">**CORRELATION MATRIX BETWEEN DATA VARIABLES**</p>


Here I'm going to explore more closely the linear relationships between variables to get a better understanding of the data. You can see the correlation coefficients of each pair from table below. It is interesting that there is almost zero correlation between SFM_edu (ratio of education, F/M) and FMlabour (ratio of labour force, F/M). Initially I thought that there may be a linear relationship between these variables at least to some extent. However, the SFM_Edu has some other interesting relationships for example with Maternal mortality ratio (mort_ratio) and Adolescent birth rate (ad_birth). These relationships are negative, suggesting that when the ratio of Female Education is high the ratios of Maternal mortality and Adolecent's birth are respectively low. You can perceive the same logic also between Expected years of schooling per country (Eedu) and the mort_ratio/ad_birth as well as between Expected life years at birth per country (Elife) and the mort_ratio/ad_birth.

Interestingly the Gross National Income per capita (GNIpc) has a positive relationship with Expected years of schooling per country (Eedu) and Expected life years at birth per country (Elife). It may be a natural consequence that if a country is wealthy they can invest more for education. But I think there might be some other components affecting to the lifespan of population even though there is a positive correlation between GNI and life years. 

If you look at the "Seats" column (or row), i.e. Share of seats in parliament (% held by women), there are no high correlation values while it has relationships with FMlabour and Eedu in to some extent. The former may suggest that when the ratio of female labour force participation is high the number of seats in parliament (held by women) can also be high and vice versa. The latter can imply that when the Expected years of schooling per country is high the number of seats in parliament (held by women) can also be high. Here I can imagine that if there are high amount of schooling years in some country there would be also more educated women among the population to be voted to the parliament. Note that the latter point of view was just "an educated guess"!

After the correlation matrix you can see a more visual "correlation matrix".



&nbsp;

```{r, message=FALSE, warning=FALSE}
library(tidyverse); library(knitr)
cor_matrix <- cor(human_own) %>% round(digits = 2)
kable(cor_matrix, format = "pandoc", digits = 2,  align="c")
```


&nbsp;

<p class="text-success">**PLOTTING THE CORRELATION MATRIX**</p>

<br>

```{r, fig.height=8, fig.width=8, message=FALSE, warning=FALSE}
library(corrplot)
corrplot(cor_matrix, method = "circle", tl.cex = 0.8, addCoef.col = "black")
```


&nbsp;


## Principal Component Analysis, PCA, with the NON-standardized data

&nbsp;

When we have a large set of correlated variables, Principal Component Analysis allows us to summarize this set with a smaller number of representative variables that collectively explain most of the variability in our dataset. PCA is an unsupervised learning algorithm which means that there is no outcome to be predicted but instead it tries to find patterns in the data. This can be done by data visualizing as seen in this and will be seen in the next chapter.

First I'm going to perform principal component analysis (PCA) without standardizing the Human data. Note that the PCA is performed again within the next chapter but with standardized dataset. Thus the purpose is to perceive and interpret if there are any differences between these methods or models and my guess is that the actual purpose is to show and justify why the data has to be standardized. 


&nbsp;

<p class="text-primary">**SUMMARY TABLE OF NON-STANDARDIZED DATA**</p>

Here you can see the same summary table of Human dataset as seen in the beginning of this chapter. It is shown again for the practical reasons to better compare the results between PCAs with non-standardized and standardized datasets.

&nbsp;

```{r}
library(knitr)
kable(summary(human_own), format = "pandoc", digits = 2,  align="l")
```


&nbsp;

<p class="text-primary">**PERFORMING PCA WITH SINGULAR VALUE DECOMPOSITION METHOD**</p>

There are two slightly different methods to perform the PCA: the Eigenvalue Decomposition and the Singular Value Decomposition. According to our instructions in DataCamp the latter is preferred.

From the tables and biplot below you can see the results of Principal Component Analysis with non-standardized Human data. The results are trying to say that there is only one principal component in the Human data. Since PCA is based on capturing the biggest variation of the dataset, you can perceive that the method captures only the variance of Gross National Income (GNI). This is because the GNI has a largest range of its scale and the other variables are not comparable with this scale. Thus, we can have such a misleading results if we forget to standardize our data. 
 


&nbsp;

```{r, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}

# Performing the PCA with Singular Value Decomposition (SVD) method
pca_human1 <- prcomp(human_own)
s1 <- summary(pca_human1)
s1

# rounded percetanges of variance captured by each PCA
pca_pr1 <- round(100*s1$importance[2, ], digits = 1)
print(pca_pr1)

# creating object pc_lab to be used as axis labels
pc_lab1 <- paste0(names(pca_pr1), " (", pca_pr1, "%)")
pc_lab1

# drawing the biplot with captions
biplot(pca_human1, cex = c(0.8, 1), col = c("dimgrey", "darkorange"), xlab = "100%", ylab = "0%")
```

&nbsp;



## Repeating the PCA with standardized data


&nbsp;

Here I provide some additional information of PCA. The purpose of PCA is to find a low-dimensional representation of a dataset that contains as much as possible of the variation. It is very important to realize that not all of these dimensions, i.e. principal components, are equally interesting. Furthermore it is known that the first principal component captures the most of the variance in the **normalized** (standardized) dataset. The second principal component captures the maximal variance of the dataset that are uncorrelated with the first component and so on.(James, Witten, Hastie & Tibshirani, 2013, 375-376.) 


&nbsp;

<p class="text-warning">**SUMMARY TABLE OF THE STANDARDIZED HUMAN DATA**</p>

As you can see from the table below, the distributions of the variables changed after stardardizing the dataset. Now the mean of every variable is zero and the observations varies on both sides of zero. This means that the different distributions are now more comparable for the further purposes.

&nbsp;


```{r}
# Standardizing the Human data
std_human <- scale(human_own)

library(knitr)
kable(summary(std_human), format = "pandoc", digits = 2,  align="l")
```

&nbsp;

<p class="text-warning">**PERFORMING PCA WITH SINGULAR VALUE DECOMPOSITION METHOD**</p>

&nbsp;

The tables and biplot below tell a huge different story about principal components in the Human data compared with earlier findings. The tables suggest that the first principal component has captured 53.6% of the total variance from the Human data. The second component has captured 16.2% and so on. As you can see, these shares are diminishing quite fast implying that the first principal components actually captures the most of the variance. 


&nbsp;

```{r, message=FALSE, warning=FALSE}

# Performing the PCA with Singular Value Decomposition (SVD) method
pca_human2 <- prcomp(std_human)
s2 <- summary(pca_human2)
s2

# rounded percetanges of variance captured by each PCA
pca_pr2 <- round(100*s2$importance[2, ], digits = 1)
print(pca_pr2)

```

&nbsp;

<p class="text-warning">**ILLUSTRATING AND INTERPRETING THE RESULTS WITH A BIPLOT**</p>

The pricipal components are illustrated in the biplot below. Just to remind that there are 155 countries and 8 variables thus the principal component score vectors have lenght n=155, and the pricipal component loading vectors have lenght p=8. Figure below plots the first two pricipal components of the data and represents both the principal component scores and the loading vectors.

From the biplot we can see that the first principal component places approximately equal weight on mort_ratio and ad_birth corresponding to their high reciprocal and positive correlation. Accordingly The first pricipal component places approximately equal weight on SFM_edu, Eedu, Elife and GNIpc which also corresponds with the high positive correlations between each of these variables. Note that each of the former variables correlate negatively with each of the latter variables". This means that these variables belong into the same principal component, i.e. representing the same phenomenon but from the opposite angles. 

Now I have to think what could be the common nominator for all of the variables listed so far. I think I will call the first principal component as the **National well-being index** since the component consists features like level and lenght of education, the lifespan and the Gross National Income combined with maternal mortality ratio and adolescent's birth rate. Let's take some examples from the biplot. Countries like Qatar, Australia and Norway seem to have a high index of wellbeing since they have high negative scores on the first component, i.e. low scores in terms of mortality and adolescent's birth rate and "high opposite scores". Accordingly countries like Chad, Niger and Sierra Leone seem to have a low index of well-being since they have high positive scores on the first component, i.e. high scores in terms of mortality and adolescent's birth rate and "low opposite scores". 

I think, in certain terms, I could compare the scale of the the first principal component to Osgood's scale which attemps to measure some polarities or contrasts but usually with opposite adjectives. But if we wanted to combine the variable for example by summing, we could translate the scale direction of either set of variables.

Then we can take a look at the other axis i.e. the second principal component. From the biplot we can see that the second component places approximately equal weight on FMlabour and seats also corresponding to their high positive correlation. Both of the variables could represent the rate of females at "work", even though the share of seats in parliament can be seen as a public "job". Nevertheless the variables could represent also some forms of equality and inequality. I think I will name the second component as **National Index of Equal Employment**. 

Let's take some examples from the biplot in terms of the second component. Countries like Iceland, Bolivia, Mozambique and especially the Rwanda have high positive scores on our "National Index of Equal Employment" while countries like Iran, Lebanon and Yemen have very low scores. 

But since the biplot is a "fourfold table", we can find more interpretations. We can start from the upper left corner where we can find Finland, earlier mentioned Iceland and Denmark, i.e. Scandinavia and some other European countries. These countries represent the ideal places considering both our indices, i.e. high in employment equality and well-being. Upper right corner has high in equality but low in well-being, e.g. Mozambigue. Lower right corner has low both in well-being and equality and there you can find for exmple Niger and Afghanistan. Finally we have lower left corner which represent countries with high in well-being but low with employment equality. There are countries like Qatar, Kuwait and Saudi-Arabia.

&nbsp;

```{r, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}

# drawing the biplot again with captions

biplot(pca_human2, cex = c(0.8, 1), col = c("dimgrey", "deeppink"), xlab = "National well-being index", ylab = "National Index of Equal Employment")
```


&nbsp;

## Multiple Correspondece Analysis, MCA

&nbsp;

Here the purpose is to get familiar with Multiple Correspondece Analysis, MCA. For this purpose I have to download a new dataset since MCA can be utilized when the data contains categorical variables and my earlier dataset contained only continuous variables. 

I'm going to download the **Tea dataset** from the **FactoMineR** packages.  

&nbsp;

<p class="text-danger">**STRUCTURE OF THE TEA DATASET**</p>

As you can imagine the data concern a questionnaire on tea consumption. There are 300 individuals who where asked how they drink tea, 18 questions. Then there are supplementary variables like age and sex. Note that the age is the only continuous variable in the dataset. 12 questions concern about individuals' product perception and 4 guestions consis some personal details.  

&nbsp;

```{r, message=FALSE, warning=FALSE}
library(FactoMineR) # https://cran.r-project.org/web/packages/FactoMineR/index.html
data("tea")
str(tea)
```


&nbsp;

<p class="text-danger">**VARIABLES TO KEEP**</p>

I'm going to reduce the number of variables and choose some specific ones:

1. breakfast
2. evening
3. lunch
4. dinner
5. Tea
6. How
7. sugar
8. how
9. where
10. price

Additionally I'm going to call this new data as **my_tea**

&nbsp;


```{r, message=FALSE, warning=FALSE}
library(dplyr)

# identifying the columns to keep
keep_columns <- c("breakfast", "evening", "lunch", "dinner", "Tea", "How", "sugar", "how", "where", "price")

# selecting the columns
my_tea <- dplyr::select(tea, one_of(keep_columns))
```

&nbsp;

<p class="text-danger">**STRUCTURE OF MY_TEA DATASET**</p>

From the table below you can perceive that there are 300 observations, i.e. individuals, and 11 variables which all are categorical. 


```{r}
str(my_tea)
```


&nbsp;

<p class="text-danger">**SUMMARY TABLE OF THE MY_TEA DATASET**</p>

From the table below you can have closer look at the categories of the kept variables. 


```{r}
summary(my_tea)
```


&nbsp;


<p class="text-danger">**VISUALIZING THE STRUCTURES OF DATASET**</p>


&nbsp;


```{r, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
library(tidyr); library(ggplot2)
gather(my_tea) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar(fill = "darkorchid1") + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```


&nbsp;

<p class="text-danger">**MULTIPLE CORRESPONDENCE ANALYSIS WITH MY_TEA DATASET**</p>


```{r}
library(FactoMineR)

# Not yet plotting
mca <- MCA(my_tea, graph = FALSE)

summary(mca)

```

&nbsp;

```{r, fig.height=8, fig.width=10}
plot(mca, invisible = c("ind"), habillage = "quali")
```



&nbsp;





