# Dimensionality reduction techniques 

&nbsp;

In this chapter the purpose is to get familiar with some of the Dimensionality Reduction Techniques including Principal Component Analysis (PCA) and Multiple Correspondece Analysis (MCA). 

Principal Component Analysis is way of identifying patterns in data, and expressing the data in such a way as to highlight their similarities and differences. PCA is a procedure After founding the patterns PCA has the tools of compressing the data by reducing the number of dimensions, without much loss of information. [Additional info of PCA](https://en.wikipedia.org/wiki/Principal_component_analysis)

Multiple Correspondence Analysis is used for the same purpose as PCA but it can be utilized if the data contains categorical variables. [Additional info of MCA](https://en.wikipedia.org/wiki/Multiple_correspondence_analysis)


&nbsp;

## Brief description of the dataset


&nbsp;

In this chapter I will be working with the **Data of Human Development approach** adopted from [United Nations Development programme, UNDP](http://hdr.undp.org/en/humandev). This programme has explored human development around the world with specific indices in each year since 1990, [Country List in 2015](http://hdr.undp.org/en/countries). The specified indices are: 

1. the Human Development Index, HDI
2. the Inequality-Adjusted Human Development Index, IHDI
3. the Gender Development Index, GDI
4. the Gender Inequality Index, GII
5. the Multidimensional Poverty Index, MPI

<br>

**I'm going to work with HDI and GII indices by combining the datasets and using specific variables including:**

1. country - technically a rowname
2. SFM_edu = SFedu / SMedu, where
    + SFedu - Females, Population with at least some secondary education (% ages 25 and older)
    + SMedu - Males, Population with at least some secondary education (% ages 25 and older)
3. FMlabour = Flabour / Mlabour, where
    + SMedu - Males, Population with at least some secondary education (% ages 25 and older)
    + Flabour - Females, Labour force participation rate (% ages 15 and older)
4. Eedu - Expected years of schooling per country
5. Elife - Expected life years at birth per country
6. GNIpc - Gross national income per capita per country in PPP, (https://en.wikipedia.org/wiki/Purchasing_power_parity)
7. mort_ratio - Maternal mortality ratio (deaths per 100,000 live births)
8. ad_birth - Adolescent birth rate (births per 1,000 women ages 15-19)
9. seats - Share of seats in parliament (% held by women)


&nbsp;

<p class="text-danger">**ACTUAL DATASET**</p>

You can find the combined dataset from my GutHub respitory: https://github.com/8475394/IODS-project/blob/master/data/human_1.txt

&nbsp;

```{r, message=FALSE, warning=FALSE}
library(tidyr); library(readr); library(knitr)
human_own <- read.table(file.path("C:/Users/heidi/Documents/YLIOPISTO/TILASTOTIEDE/INTRODUCTION TO OPEN DATA SCIENCE/IODS-project/data", "human_1.txt"), header = TRUE)
```

&nbsp;

<p class="text-danger">**STRUCTURE OFTHE HUMAN DATASET**</p>

From the structure table you can notice that there are 155 observation and 8 variables. Note that the country is a rowname and it doesn't appear to variable list.

&nbsp;

```{r, message=FALSE, warning=FALSE}
library(knitr)
table(str(human_own))
```

&nbsp;

## Summaries and graphs


&nbsp;

```{r, results='asis'}
library(knitr)
kable(summary(human_own), format = "pandoc", caption = "Summaries of Human Data", digits = 2,  align="l")

```

&nbsp;



```{r}
library(ggplot2); library(gridExtra);library(dygraphs)

# INITIALIZING THE PLOTS

qplot(human_own$SFM_edu, geom = "histogram", binwidth = 0.1, main = "kokeilu", xlab = "education", col = I("grey"), fill = I("chartreuse3"))

qplot(human_own$FMlabour, geom = "histogram", binwidth = 0.08, main = "kokeilu", xlab = "kokeilu", col = I("grey"), fill = I("orange"))

qplot(human_own$Eedu, geom = "histogram", binwidth = 1, main = "Expected years of schooling per country", xlab = "schooling", col = I("grey"), fill = I("mediumorchid"))
```


&nbsp;

